### 2022-12-27 paper.Rmd
round of editing

    Anticlustering is a collection of methods to partition data sets into disjoint groups while aiming for high between-group similarity or high within-group heterogeneity. Anticlustering thereby reverses the logic of its better known twin---cluster analysis---and is usually approached by maximzing instead of minimizing a clustering objective function. Introducing k-plus, this paper presents an extension of the k-means clustering objective to maximize between-group similarity in anticlustering applications. The k-plus criterion represents between-group similarity as discrepancy in distribution moments (means, variance, higher order moments, and covariances), thereby extending the classical k-means criterion, which only reflects how groups differ in their means. It is shown that the k-plus criterion can be reduced to an augmentation of the input data while the original k-means objective remains unchanged. A computer simulation and practical examples show that k-plus anticlustering achieves high between-group similarity with regard to multiple objectives. In particular, optimizing between-group similarity with regard to variances as well as means comes with almost no cost to either criterion; the k-plus extension is therefore generally preferred over classical k-means anticlustering. Higher order moments and covariance structure can also succesfully be equalized. Example are given on how k-plus anticlustering can be applied to real norming data using the open source R package `anticlust`, which is freely available via CRAN (https://cran.r-project.org/package=anticlust) and Github (https://github.com/m-Py/anticlust).

@@

Anticlustering is a collection of methods to partition elements into disjoint groups, with the goal of obtaining high between-group similarity or within-group heterogeneity [@brusco2019; @spath1986; @valev1998]. Oftentimes, both goals coincide: some anticlustering objectives imply that similarity between groups is maximal whenever similarity within groups is minimal [@feo1990]. Anticlustering thereby reverses the logic of its better known twin cluster analysis, which seeks homogeneity within clusters and heterogeneity between clusters [@rokach2005; @steinley2006]. Whereas a variety of different approaches exist for cluster analysis---including hierarchical methods, model-based clustering, or network approaches [see @brusco_emergent_2012, for an overview]---anticlustering is usually formalized as a partitioning process that optimizes an objective function that represents between-group similarity and/or within-group heterogeneity [@papenberg2020; @spath1986; @brusco2019].

Anticlustering is useful in many tasks surrounding research work in psychology [@steinley2006]. For example, anticlustering can be used to split tests into parts of equal difficulty [@gierl2017], to assign students to work groups [@baker2002] or when assigning stimuli to different---but parallel---experimental conditions [@lahl2006]. For a more comprehensive overview of anticlustering applications in Psychology, see @brusco2019 and @papenberg2020. Solving anticlustering problems "by hand" is a tedious and time-consuming task, and the quality of manual partitioning is usually subpar. Fortunately, these problems can be formalized as mathematical optimization problems [e.g., @brusco2019; @spath1986; @baker2002; @fernandez2013] and accessible open source software solutions to tackling these problems exist [@papenberg2020].

Anticlustering methods can be distinguished on the basis of the input data and the objective function that is used to quantify between-group similarity or within-group heterogeneity. Some anticlustering objectives are based on pairwise dissimilarity ratings such as the Euclidean distance. The most prominent criterion of this kind is the *diversity*, which is the total sum of dissimilarities between any elements within the same group [@brusco2019]. By summing dissimilarities within-groups, the diversity is primarily a measure of within-group heterogeneity. For equal-sized groups, however, maximizing the diversity simultaneously minimizes the average dissimilarity between elements in different groups, thereby maximizing between-group similarity.

Another important anticlustering criterion that is based on pairwise dissimilarities is the *dispersion*, i.e., the minimum dissimilarity between any two elements within the same group [@fernandez2013]. Maximizing the dispersion increases the within-group heterogeneity by ensuring that any two elements in the same group as dissimilar from each other as possible. @brusco2019 convincingly argued that anticlustering applications requiring high within-group heterogeneity should incorporate both dispersion and diversity, presenting a bicriterion algorithm to approximate the Pareto efficient set of solutions according to both criteria.

Oftentimes, partitioning applications in psychology use attribute data [or *features*; @dry2009] instead of dissimilarity ratings as input. This use case is for example common when selecting stimuli for an experiment on the basis of norming data [e.g., @kurdi2017introducing]. For word stimuli in psycholinguistic experiments, norming data may consist of ratings for imagery and concreteness, as well as orthographic variables [@friendly1982]. In other cases, attribute values are binary and may represent the presence or absence of a feature [@tversky1977]. In the context of anticlustering, one approach for handling attribute data is to compute an appropriate dissimilarity measure such as the pairwise Euclidean or squared Euclidean distance across the set of attributes [@brusco2019]. Subsequently, an anticlustering criterion such as the diversity can be optimized to obtain a partitioning of the data set. A different anticlusting approach is based on clustering objectives that can directly be computed on the attribute values. K-means is probably the best-known method of this kind. In k-means clustering, the sum of the squared Euclidean distances between data points and their cluster centers is minimized [@brusco2006branch], usually using the k-means heuristic [@jain2010; @steinley2006]. Minimizing the k-means criterion simultaneously maximizes between-cluster separation and within-cluster homogeneity [@aloise2009np]. Reversing the k-means objective function---using maximization instead of minimization---has been recognized as a useful approach when aiming for high between-group similarity [@spath1986; @valev1998; @valev1983]. Specifically, as @spath1986 noted, k-means anticlustering minimizes differences with regard to the means of the numeric attributes across clusters. However, other parameters---such as the variance---are ignored. @papenberg2020 showed that k-means anticlustering tends to over-optimize similarity with regard to the mean; even an entirely random split usually resulted in more similar variances. When requiring that groups are overall similar to each other, k-means' focus may be misguided, because similar means can be obtained even when the underlying distributions are clearly different [e.g., @anscombe1973].

To optimize overall between-group similarity---and hence to overcome the limitations of k-means anticlustering---this paper introduces k-plus anticlustering. K-plus is an extension of the k-means objective that includes one or several terms to quantify between-group similarity with regard to other distribution characteristics instead of only the mean. Specifically, k-plus offers the possibility to minimize differences with regard to the variance, covariances, and higher order moments such as skewness and kurtosis. The k-plus criterion is not a single criterion for anticlustering, but a familiy of objectives implementing different requirements with regard to between-group similarity. After formally introducing the k-plus objective, a simulation study is reported showing that k-plus anticlustering is well suited to create groups having minimal differences according to multiple distribution characteristics. Two practical examples are included that illustrate how readers can easily apply k-plus anticlustering using the free and open source software package `anticlust`, an extension to the popular statistical programming language R [@papenberg2020; @R-base; @R-anticlust].

@@

Because the sample variance is basically defined as an average, k-means anticlustering can be employed to equalize the variance across groups. That is, we can formulate an adaption of the $\mathit{SSE}$ to quantify the degree to which the variance differs between groups. First, however, we generalize the idea to the case of $P$ attributes. For each attribute, we have to compute the squared deviation between all values and the mean of the attribute. We obtain a new data matrix $\mathbf{Y}_{N \times P} = \{y_{ij}\}_{N \times P}$, where

@@

After the five anticlustering methods and the random allocation procedure had been applied to the 10,000 data sets for $K = 2$, $K = 3$, $K = 4$, I investigated how well the competing methods were able to create similar groups. To this end, I investigated the discrepancy in means, standard deviations, skewness, kurtosis, and correlation between groups---for each data set for each $K$. To quantify the discrepancy in group means, the following computation was used: for each of the (2, 3, 4 or 5) features, all (2, 3 or 4) group means were computed. For each feature, the difference between the minimum and maximum group mean was used as a measure of discrepancy; the global discrepancy in means ($\Delta M$) was then determined as the average discrepancy across features. The same procedure was applied to compute the discrepancy in standard deviations ($\Delta \mathit{SD}$), skewness ($\Delta \mathit{Skew}$) and kurtosis ($\Delta \mathit{Kurtosis}$). Quantifying discrepancy in correlations ($\Delta \mathit{Cor}$) followed the same rule, but discrepancies in correlations between groups had to be aggregated across pairs of features instead of single features.

@@

Table 1 displays the global simulation results aggregated across all conditions. Standard k-means performed best at minimizing discrepancy with regard to means ($\Delta \mathit{M}$), but did not outperform a random assignment with regard to all other distribution characteristics; replicating @papenberg2020, k-means' performance with regard to similarity in standard deviations was even below that of a random assignment. K-plus anticlustering addressed $\Delta \mathit{M}$ almost as well as k-means, and at the same time minimized discrepancy with regard to the standard deviations ($\Delta \mathit{SD}$); all three k-plus approaches outperformed all other methods with regard to minimizing $\Delta \mathit{SD}$.

As Figure 1 shows, with increasing $N$, k-means and all k-plus methods converge on minimizing $\Delta \mathit{M}$. This is an important observation because it shows that including the k-plus term in the optimization process does not necessarily prevent minimizing discrepancy in means; instead, multiple objectives can be addressed at the same time---especially when $N$ increases. However, while k-plus anticlustering is capable of addressing multiple objectives simultaneously, increasing the number of optimization criteria apparently aggravates fulfilling each single criterion. K-plus skew/kurtosis and k-plus correlation performed worse at minimizing discrepancy with regard to standard deviations than standard k-plus, which itself performed slightly worse than k-means anticlustering. Still, every k-plus method did what it was supposed to do: k-plus skew/kurtosis was best at minimizing $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$, k-plus correlation was best at minimizing discrepancy with regard to $\Delta \mathit{Cor}$. At the same time, these two k-plus objectives maintained a comparably good level of addressing $\Delta \mathit{M}$ and $\Delta \mathit{SD}$, outperforming diversity anticlustering.

@@

Generally, it should be noted that diversity anticlustering is a good all-arounder method that tends to somewhat addressed all distribution characteristics. For the specific objectives that are addressed by the k-plus methods, however, diversity anticlustering was outperformed. As as side note, apart from k-plus correlation, diversity anticlustering was the only method that equalizes correlation structure among groups. To the best of my knowledge, this observation has not been made previously, even though the maximum diversity problem has been widely studied.

@@

The simulation study indicated that k-plus may show decreased performance for smaller group sizes. Therefore, a second application illustrates that k-plus anticlustering yields satisfactory results in practice even when group sizes are smaller. I make use of a norming data set of 96 word stimuli that was contributed to the `anticlust` package by Dr. Schaper [@schaper2019metacognitive; @schaper2019metamory; also see @papenberg2020 for a description of the norming data]. After loading the `anticlust` package, the data set can be accessed as follows:

@@

apa_table(
  tab,
  caption = "Descriptive statistics by group for the Schaper et al. (2019a; 2019b) data set, after applying default k-plus anticlustering.",
  note = "The cells contain mean values by cluster and the standard deviation (in parentheses).",
  escape = FALSE,
  align = c("r")
)

@@

The present paper focused on the new k-plus criterion for anticlustering, but only superficially considered the multicriterion nature of this criterion. Instead of applying a specialized multicriterion optimization algorithm, a simple (unweighted) sum approach---combined with a generic local maximum neighbourhood search---was used to optimize the k-plus objective. The same approach was used independent of the number of objectives that were included as part of the anticlustering process. More sophisticated algorithmic approaches may yield improved results, as well as the opportunity to deliberately select among partitionings that may be preferred according to different k-plus objectives. @brusco2019 showed how direct algorithms can be used to optimize multiple criteria at the same time in anticlustering applications. Their study focused on approximating the pareto set of efficient solutions that optimize diversity and dispersion, which is essential in applications that require high within-group heterogeneity. Future research should investigate if k-plus anticlustering may profit from similar algorithmic treatment.

This paper's focus on the k-plus objective only instead of optimization algorithms followed two reasonings. First, the applications showed that the local maximum search---maximizing the unweighted k-plus sum---yields satisfying results in practice that probably suffice most applications. Second, it was of interest to investigate the merits of the k-plus objective function per se, independently from the algorithm that is used to optimize it. To obtain a fair assessment In the simulation study, it was therefore important that each anticlustering objective is optimized using the same algorithm, instead of using specialized optimization schemes tailored to each individual criterion. It was of particular interest to compare the k-plus objective with the popular diversity objective. @papenberg2020 compared k-means anticlustering and diversity anticlustering[^clusterediting] in a simulation study and concluded that the diversity criterion should be preferred because k-means does not equalize the spread of the distribution between groups. In contrast, maximizing the diversity leads to an appropriate balance between location and spread. Using the k-plus extension, however, the k-means criterion has regained attractivity: When opting for similarity in means and variances---and for increasing $N$, skewness and kurtosis as well---, k-plus anticlustering clearly outperformed diversity anticlustering.

>>>>

    Anticlustering refers to the process of partitioning elements into disjoint groups with the goal of obtaining high between-group similarity and high within-group heterogeneity. Anticlustering thereby reverses the logic of its better known twin---cluster analysis---and is usually approached by maximzing instead of minimizing a clustering objective function. Introducing k-plus, this paper presents an extension of the k-means clustering objective to maximize between-group similarity in anticlustering applications. The k-plus criterion represents between-group similarity as discrepancy in distribution moments (means, variance, higher order moments, and covariances), thereby extending the classical k-means criterion, which only reflects how groups differ in their means. It is shown that the k-plus criterion can be reduced to an augmentation of the input data while the original k-means objective remains unchanged. A computer simulation and practical examples show that k-plus anticlustering achieves high between-group similarity with regard to multiple objectives. In particular, optimizing between-group similarity with regard to variances as well as means comes with almost no cost to either criterion; the k-plus extension is therefore generally preferred over classical k-means anticlustering. Higher order moments and covariance structure can also succesfully be equalized. Example are given on how k-plus anticlustering can be applied to real norming data using the open source R package `anticlust`, which is freely available via CRAN (https://cran.r-project.org/package=anticlust) and Github (https://github.com/m-Py/anticlust).

@@

Anticlustering refers to the process of partitioning elements into disjoint groups with the goal of obtaining high between-group similarity and high within-group heterogeneity [@brusco2019; @spath1986; @valev1998]. Anticlustering thereby reverses the logic of its better known twin---cluster analysis---which seeks homogeneity within clusters and separation between clusters [@rokach2005; @steinley2006]. Anticlustering has many applications in research psychology [@steinley2006; @brusco2019; @papenberg2020]: Examples include splitting tests into parts of equal difficulty [@gierl2017], assigning students to work groups [@baker2002], and assigning stimuli to different, but parallel, experimental conditions [@lahl2006]. Solving anticlustering problems "by hand" is a tedious and time-consuming task, and the quality of manual partitioning is usually subpar. Fortunately, these problems can be formalized as mathematical optimization problems [e.g., @brusco2019; @spath1986; @baker2002; @fernandez2013] and accessible open source software solutions to tackling these problems exist [@papenberg2020].

Anticlustering methods are characterized by (a) an objective function that quantifies between-group similarity and/or within-group heterogeneity and (b) an algorithm that determines how elements are assigned to groups in such a way that the objective function is maximized. Several anticlustering objective functions use pairwise dissimilarity ratings such as the Euclidean distance as input. The most prominent criterion of this kind is the *diversity*, which is the total sum of dissimilarities between any elements within the same group [@brusco2019; @gallego2013]. By summing all pairwise dissimilarities of elements within the same group, the diversity reflects the total degree of within-group heterogeneity. High within-group diversity simultaneously ensures high similarity of the different groups.[^equalsizedgroups] Another important anticlustering criterion is the *dispersion*, which is the minimum dissimilarity between any two elements within the same group [@fernandez2013]. Maximizing the dispersion increases the within-group heterogeneity by ensuring that any two elements in the same group as dissimilar from each other as possible. @brusco2019 convincingly argued that anticlustering applications requiring high within-group heterogeneity should incorporate both dispersion and diversity, presenting a bicriterion algorithm to approximate the Pareto efficient set of solutions according to both criteria.

[^equalsizedgroups]: As will be discussed later, the diversity objective only stricly equalizes within-group heterogeneity and between-group similarity when groups are equal-sized.

Oftentimes, partitioning applications in psychology are focused on ensuring between-group similarity rather than within-group heterogeneity (even though both goals oftentimes coincide). High similarity between-group similarity is for example desireable when designing experimental conditions that have to incorporate different stimulus sets that should however be as similar as possible according to response-relevant attributes [@lahl2006]. Such stimulus assignment tasks are usually conducted on the basis of attribute data [or *features*; @dry2009] rather than pairwise dissimilarity ratings, for example when selecting stimuli for an experiment on the basis of norming data [e.g., @kurdi2017introducing]. For word stimuli in psycholinguistic experiments, norming data may consist of ratings for imagery and concreteness, as well as orthographic variables [@friendly1982]. In other cases, attribute values are binary and may represent the presence or absence of a feature [@tversky1977]. In the context of anticlustering, one approach for handling attribute data is to compute an appropriate dissimilarity measure such as the pairwise Euclidean or squared Euclidean distance across the set of attributes [@brusco2019]. Subsequently, an anticlustering criterion such as the diversity can be optimized to obtain a partitioning of the stimuli. A different anticlusting approach employs clustering objectives that can directly be computed on the attribute values. K-means is probably the best-known objective of this kind. In k-means clustering, the sum of the squared Euclidean distances between data points and their cluster centers is minimized [@brusco2006branch], usually using the k-means heuristic [@jain2010; @steinley2006]. Minimizing the k-means criterion simultaneously maximizes between-cluster separation and within-cluster homogeneity [@aloise2009np]. Reversing the k-means objective function---using maximization instead of minimization---has been recognized as a useful approach when aiming for high between-group similarity [@spath1986; @valev1998; @valev1983]. Specifically, as @spath1986 noted, k-means anticlustering minimizes differences with regard to the means of the numeric attributes across clusters. However, other parameters---such as the variance---are not targeted. @papenberg2020 showed that k-means anticlustering tends to over-optimize similarity with regard to the mean; even an entirely random group assignment usually yielded more similar variances. When requiring overall between-group similarity, the negligence of other distribution characteristics than the mean is misguided, because similar means can be obtained even when the underlying distributions are clearly different [e.g., @anscombe1973].

To optimize overall between-group similarity---and hence to overcome the limitations of k-means anticlustering---this paper introduces k-plus anticlustering. K-plus is an extension of the k-means objective that quantifies between-group similarity as discrepancy with regard to other distribution characteristics instead of only the mean. Specifically, k-plus offers the possibility to minimize differences with regard to the variance, covariances, and higher order moments such as skewness and kurtosis. After formally introducing the k-plus objective, a simulation study is reported showing that k-plus anticlustering is well suited to create groups having minimal differences according to multiple distribution characteristics. Practical examples illustrate how readers can easily apply k-plus anticlustering using the free and open source software package `anticlust`, an extension to the popular statistical programming language R [@papenberg2020; @R-base; @R-anticlust].

@@

Because the sample variance is basically defined as an average, k-means anticlustering can be employed to equalize the variance across groups, which is the central insight motivating the new k-plus criterion. To obtain an adaption of the $\mathit{SSE}$ that quantifies the degree to which the variance differs between groups, we have to compute the squared deviation between all individual values and the mean of the attribute. We obtain a new data matrix $\mathbf{Y}_{N \times P} = \{y_{ij}\}_{N \times P}$, where

@@

After the five anticlustering methods and the random allocation procedure had been applied to the 10,000 data sets for $K = 2$, $K = 3$, $K = 4$, I investigated how well the competing methods were able to create similar groups. To this end, I investigated the discrepancy in means, standard deviations, skewness, kurtosis, and correlation between groups---for each data set for each $K$. To quantify the discrepancy in group means, the following computation was used: for each of the (2, 3, 4 or 5) features, all (2, 3 or 4) group means were computed. For each feature, the difference between the minimum and maximum group mean was used as a measure of discrepancy; the global discrepancy in means ($\Delta M$) was then determined as the average discrepancy across features. The same procedure was applied to compute the discrepancy in standard deviations ($\Delta \mathit{SD}$), skewness ($\Delta \mathit{Skew}$) and kurtosis ($\Delta \mathit{Kurtosis}$). Quantifying discrepancy in correlations ($\Delta \mathit{Cor}$) followed the same rule, but discrepancies in correlations between groups had to be averaged across pairs of features instead of single features.

@@

Table 1 displays the global simulation results aggregated across all conditions. Standard k-means performed best at minimizing discrepancy with regard to means ($\Delta \mathit{M}$), but did not outperform a random assignment with regard to all other distribution characteristics. Replicating @papenberg2020, k-means' performance with regard to similarity in standard deviations was even below that of a random assignment. K-plus anticlustering addressed $\Delta \mathit{M}$ almost as well as k-means, and at the same time minimized discrepancy with regard to the standard deviations ($\Delta \mathit{SD}$); all three k-plus approaches outperformed all other methods with regard to minimizing $\Delta \mathit{SD}$.

As Figure 1 shows, with increasing $N$, k-means and the k-plus methods converge on the same level with regard to minimizing $\Delta \mathit{M}$. This is an important observation because it shows that including the k-plus term in the optimization process does not necessarily prevent minimizing discrepancy in means; instead, multiple objectives can be addressed at the same time---especially when $N$ increases. However, while k-plus anticlustering is capable of addressing multiple objectives simultaneously, increasing the number of optimization criteria apparently aggravates fulfilling each single criterion. K-plus skew/kurtosis and k-plus correlation performed worse at minimizing discrepancy with regard to standard deviations than standard k-plus, which itself performed slightly worse than k-means anticlustering. Still, every k-plus method did what it was supposed to do: k-plus skew/kurtosis was best at minimizing $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$, k-plus correlation was best at minimizing discrepancy with regard to $\Delta \mathit{Cor}$. At the same time, these two k-plus objectives maintained a comparably good level of addressing $\Delta \mathit{M}$ and $\Delta \mathit{SD}$, outperforming diversity anticlustering.

@@

Generally, it should be noted that diversity anticlustering is a good all-arounder method that tends to somewhat addressed all distribution characteristics. For the specific objectives that are addressed by the k-plus methods, however, diversity anticlustering was outperformed. As as side note, apart from k-plus correlation, diversity anticlustering was the only method that equalized correlation structure among groups. To the best of my knowledge, this observation has not been made previously, even though the maximum diversity problem has been widely studied.

@@

The simulation study indicated that k-plus may show decreased performance for smaller group sizes. Therefore, a second application illustrates that k-plus anticlustering yields satisfactory results in practice even when group sizes are smaller. I make use of a norming data set of 96 word stimuli that was contributed to the `anticlust` package by Marie L. Schaper [@schaper2019metacognitive; @schaper2019metamory; also see @papenberg2020 for a description of the norming data]. After loading the `anticlust` package, the data set can be accessed as follows:

@@

apa_table(
  tab,
  caption = "Descriptive statistics by group for the Schaper et al. (2019a; 2019b) data set, after applying k-plus anticlustering.",
  note = "The cells contain mean values by cluster and the standard deviation (in parentheses).",
  escape = FALSE,
  align = c("r")
)

@@

The present paper focused on the new k-plus criterion for anticlustering, but only superficially considered the multicriterion nature of this criterion: A simple (unweighted) sum approach---combined with a generic local maximum neighbourhood search---was used to optimize the k-plus objective. More sophisticated multicriterion approaches may yield improved results, as well as the opportunity to deliberately select among partitionings that may be preferred according to different k-plus objectives (means, variances, higher order moments, and covariances). @brusco2019 showed how direct algorithms can be used to optimize multiple criteria at the same time in anticlustering applications. Their study focused on approximating the pareto set of efficient solutions that optimize diversity and dispersion, which is essential in applications that require high within-group heterogeneity. Future research should investigate if k-plus anticlustering may profit from similar algorithmic treatment.

This paper's focus on the k-plus objective only instead of optimization algorithms followed two reasonings. First, the applications showed that the local maximum search---maximizing the unweighted k-plus sum---yields satisfying results in practice that probably suffice in most applications. Second, it was of interest to investigate the merits of the k-plus objective function per se, independently from the algorithm that is used to optimize it. To obtain a fair assessment In the simulation study, it was therefore important that each anticlustering objective is optimized using the same algorithm, instead of using specialized optimization schemes tailored to each individual criterion. It was of particular interest to compare the k-plus objective with the popular diversity objective. @papenberg2020 compared k-means anticlustering and diversity anticlustering[^clusterediting] in a simulation study and concluded that the diversity criterion should be preferred because k-means does not equalize the spread of the distribution between groups. In contrast, maximizing the diversity leads to an appropriate balance between location and spread. Using the k-plus extension, however, the k-means criterion has regained attractivity: When opting for similarity in means and variances---and for increasing $N$, skewness and kurtosis as well---, k-plus anticlustering clearly outperformed diversity anticlustering.
